My advisor David Bindel has my deepest appreciation, for this work is only
possible with his guidance. I thank him not only for sharing his vast knowledge
and experience, but also for the patience, gentleness, and encouragement in his
mentorship. I will forever carry the lessons I learned from him.

I want to thank my committee members Andrew Wilson and Adrian Lewis, as well as
my temporary advisor Shane Henderson. Andrew led me into an important area of my
research, which makes up Chapter 3 of this thesis. Adrian and Shane both offered
my valuable advice in research and life as a graduate student.

I deeply appreciate all my wonderful collaborators. Chapter 2 was co-authored
with Austin Benson and David Bindel. Chapter 3 was co-authored with David
Bindel, David Eriksson, Eric Lee, Hannes Nickisch and Andrew Wilson. Chapter 4
was co-authored with David Bindel, Sungjun Cho, Moontae Lee, and David Mimno.
Chapter 5 was co-authored with Wei Hu and Lin Lin. Thank you for the time and
effort in working with me. It was truly a pleasure to collaborate with you all.

I also want to thank all the students and staffs at CAM, especially  Andrew
Loeb, Zachary Clawson, Eric Lee, David Eriksson, and my best friend Marc Aur\`
{e}le Tiberius Gilles. You made this part of my life not just rewarding, but
enjoyable as well.

I would like to acknowledge my colleagues during my internships at Lawrence
Berkeley National Laboratory and Google Network Infrastructure Team,
particularly my mentors Lin Lin and Rui Wang. You taught me many useful skills,
and helped me become a better researcher and programmer.

Most importantly, I am sincerely grateful for my supportive family and wife. You
have always been by my side even when I am far away.

My work is supported by the Cornell Center for Applied Mathematics, the National
Science Foundation Grant DMS-1620038, teaching assistantship in Cornell's
Department of Computer Science and Department of Mathematics, and the Cornell
University Fellowship. During the summer of 2017, I am supported by the Lawrence
Berkeley National Laboratory as a graduate research intern.