Kun Dong was born in Shaoxing, Zhejiang Province, China on September 26, 1992 to
Jian Dong and Chamei Sang. Kun became interest in puzzles and mathematics in
early elementary school, and participated in mathematics competitions until the
end of high school.

In 2008, Kun moved to Newmarket, Ontario, Canada and attended Sir William Mulock
Secondary School. He spent the next two years learning the new language and
culture.

After graduation, Kun attended the University of California, Los Angeles in 2010
to study applied mathematics. During the summers of 2012 and 2013, he
participated in the UCLA Applied Math REU program to work on dynamical system
and crime modeling. He received tremendous help and encouragement on his way to
graduate school from his REU mentors, Scott McCalla and James von Brecht. In
2014, Kun completed the departmental scholar program in mathematics, earning his
B.S. and M.A. concurrently. He graduated Summa Cum Laude, receiving the
departmental highest honor in applied mathematics and the Daus memorial award
for his achievement in mathematics as an undergraduate.

In 2014, Kun was admitted to the Ph.D. program at Center for Applied Mathematics
in Cornell University. He soon started working with Professor David Bindel, who
later became his advisor. During the summer of 2017, he interned at the Lawrence
Berkeley National Laboratory, where he worked under the mentorship of Professor
Lin Lin. For the last three years at Cornell, he was partially supported by a
National Science Foundation grant. In September 2019, Kun will move to Seattle
and become a research scientist at Facebook, Inc.